ASTR480 Final Report:\\ \textit{Asteroids in TESS}

### 2024-10-19 FinalReport/FinalReportDraft.tex
?final? Figures added from Ozstar

\autoref{Fig:FreqVsDiam} has been constructed from data from the Light Curve Database (LCDB) \citet{Warner2009}, as cited in text. That data has being built suplimented with my own work in the corrseponding Figure %TODO

@@

Asteroids can have their rotation properties changed with time, collisions can do this, but so can light from the sun.
The YORP effect (named for the discoverers Yarkovsky, O'Keefe, Radzievskii, and Paddack) is the process of tourquing asteroids of an asymmetric shape by both bombardment of photons and thermal (re-)emission \citep{Rubincam2000}.

The data on solar system bodies will only increase with more large survey telescopes coming online soon, such as the Legacy Survey of Space and Time \citep[LSST][]{LSST2019}.
The number of new asteroids that are well characterised will increase by more than 2 orders of magnitude on the Sloan Digital Sky Survey \citep[SDSS][]{York2000} which had $\sim 88000$ objects \citet{Parker2008}.
LSST needs a dedicated pipeline for asteroid classification, the Solar System Notification Alert Processing System \citep[SNAPS][]{Trilling2023} has been developed using ZTF data and is can scale up handle volume of data coming from LSST.
Work is being done to understand the biases of how  asymmetric shape high amplitude variations can lead to selection effects of such large sample of asteroids \citet{Levine2023}.

Because of such a rich history of study, many asteroids have a known rotation period.
These are catalogued by \citet{Warner2009} %TODO updated in citation http://www.MinorPlanet.info/php/lcdb.php
in the Light Curve Database (LCDB).
In the October \nth{1}, 2023 release, the LCDB has nearly 9000 objects with periods determined to a quality code $U\geq 2-$, which is the minimum reliability recommended by \citeauthor{Warner2009} for statistical analysis.
This includes other TESS data from \citet{Pal2020} and \citet{Woods2021}.
The frequencies against diameter plot of these asteroids in \autoref{Fig:FreqVsDiam} shows the collection of all the asteroids in the LCDB.
The diameter presented in the LCDB here calculated from the absolute magnitude $H$ of the bodies, assuming an albedo.
\citeauthor{Warner2009} acknowledge that there are direct measurements of diameter for more asteroids in the present day, but continue to present the $H$ derived diameters for consistency.

The asteroid spin barrier \citep{Pravec2000} is seen in the red dashed lines in \autoref{Fig:FreqVsDiam}.
There are very few asteroids larger than $\qty{0.15}{\kilo\metre}$ rotating faster than $\sim \qty{10}{\per\day}$.
It is thought that the YORP effect, as wel as collisons, spin up asteroids so fast that they pass above this barrier and then no longer have the internal strength necessary to hold themselves together.
This implies that all asteroids made to spin faster have torn themselves apart and that they have no tensile strength, i.e they are rubble piles instead of monoliths \citep{Harris1996}.
This idea of the cohesive strength of small bodies is still explored today \citep[e.g.][for V-type asteroids]{Oszkiewicz2020}.

Comparing the periods I find of asteroids that are already in the LCDB is key to making sure my methods are reliable, and that the periods for asteroids for which no data exists can be trusted.

\begin{figure}[t!]
  \centering
  \includegraphics[width=0.7\textwidth]{../Test Code/Testing Figures/Diam-FreqPlot.pdf}
  \caption[Asteroid frequencies vs diameter (LCDB)]{
    Rotaion rate of asteroids by their size, recreated from the  October \nth{1}, 2023 LCDB after \citet{Warner2009}.
  }
  \label{Fig:FreqVsDiam}
\end{figure}

@@

There is ongoing analysis of subsets of the asteroid population in TESS.
\citet{Gowanlock2024} uses TESS photometry as well as  observations by the Zwicky Transient Facility \citep[ZTF, ][]{Bellm2018} to get a longer baseline on mutually observed objects while combining ground and space based observations.
There were not very many objects in both samples with known periods for comparison, with only 222 objects analysed, however they demonstrate that the technique is effective.
Fainter and unknown solar system objects can be found by shift and stacking \citep{Holman2019, Payne2019,Rice2020} or taking a fast X-ray transform \citep{Nguyen2024} of the FFI data.

This work aims to extend these studies to more sectors, hence characterising more asteroids.
The aim is to get a volume complete set of asteroid lightcurves and periods, by analysis of the whole sky as seen by TESS using data from more than just the first year of TESS's operation.
This has not been achieved, as only one sectors analysis is presented here, but the methods are scalable.
Because of the survey properties, TESS provides a self-consistent way to measure the properties of asteroids over the full sky.
By using the higher cadence FFIs of later TESS sectors, the $P\leq\qty{3}{\hour}$ limit on the accurate periods \citep[as found by ][]{McNeill2023} can be lowered.
I also employ a different data reduction method, reducing the data with the \texttt{TESSreduce} package \citep{Ridden-Harper2021} should increase the accuracy of the photometry due to the improved background subtraction process.

Another beneficial part of this work is that as part of a full sky transient survey using TESS, \texttt{TESSELLATE} \citep{TESSELLATE}.
Asteroids can spike the brightness of a pixel for only a few frames while it is passing by.
This can confuse pipelines looking for  transients such as flare stars and supernova, as they can look very similar, having the brightness change in one pixel for a short time.
Knowing where the asteroids are will allow for the removal of these spikes from the transient pipeline, as well as to understand the asteroid population better.
By filtering the asteroid detections out, while also self consistently determining rotation properties of a large sample of the population, more science can be done with the one set of analyses.

@@

The \texttt{TESSELLATE} pipeline has been running on the OzSTAR supercomputing facilities.
I built the asteroid detection and subsequent lightcurve and periodogram analysis, to be described below, by testing on one subsection of one CCD of one camera  of one TESS sector.
Once I was confident this worked as requird, the same code was refactored slightly to work on OzSTAR and a large-scale analysis of all the processed TESS sectors can be run.
I will be looking for completeness of detections of asteroids, as well as accuracy of periods and amplitude variation.

For this work, I present analysis of TESS sector 29\footnote{\url{https://tess.mit.edu/observations/sector-29/}}, which was observed between \nth{26} August and \nth{22} September 2020.
This was the third TESS sector to be using \qty{10}{\min} cadence FFIs, which has had less analysis of the asteroids preformed.
It was chosen mainly because of this higher cadence, it was the only sector of this cadence that had been processed by \texttt{TESSELLATE} with enough time for my analysis to be preformed.
What is presented for this section is scalable, on a per sector basis, to any sector.

@@

To check for asteroids in the TESS data, the positions of the asteroids with time are required.
For most asteroids, their orbital elements are well known, so it is a matter of looking them up and cross-matching with transients in the TESS data.
Python was used to make API calls to
%{Skybot}\footnote{\href{https://vo.imcce.fr/webservices/skybot/}{Skybot: https://vo.imcce.fr/webservices/skybot/}} 
Skybot \citep{Berthier2006} to get positions of asteroids in a cone search box in right ascension (RA) and  declination (Dec) space.
The search was limited to only asteroids with an average visual magnitude $V\geq \qty{20}{\mag}$.
As TESS sectors are \qty{27}{\day} long, so querying every \qty{12}{\hour} is both computationally feasible and allows for a good understanding of a bodies motion. %TODO cite numpy
These positions are still very sparsely spaced in time compared to the TESS data, so an interpolation is needed to bridge the gap.

Knowing where the asteroids are is helpful for finding them in the archival TESS data, but knowing more about the individual asteroids is also useful for population statistics.
The \texttt{astroquery} \citep{Ginsburg2019} package was used to query JPL Horizons% {JPL Horizons}\footnote{\href{https://ssd.jpl.nasa.gov/horizons/}{JPL Horizons: https://ssd.jpl.nasa.gov/horizons/}} 
was used to obtain the orbital elements of each asteroid, $a$, $e$, and $i$, as well as an absolute magnitude $H$, which is a good proxy for size (as discussed above). %TODO ref

@@

\begin{figure}[h!]
  \begin{subfigure}[t]{0.5\textwidth}
    \centering
    \includegraphics[width =\textwidth]{../Test Code/Testing Figures/interpPos\_22\_2\_3\_4.pdf}
    \caption[Positions]{Positions of asteroids in a subsection.}
    \label{Fig:interpPos}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.5\textwidth}
    \centering
    \includegraphics[width =\textwidth]{../Test Code/Testing Figures/1989 BC1PosCheck.pdf} %TODO one with more points???
    \caption[Errors]{Error in interpolated positions}
    \label{Fig:errPos}
  \end{subfigure}

  \caption[Interpolated positions and error of asteroids]{
    (a)
    The interpolated positions of asteroids in one subsection of a TESS sector.
    The colours of the lines are time sequenced, as shown in the colour bar.
    The alpha of the colours are scaled by the absolute magnitude $H$ of the asteroid, queried from JPL Horizons.
    Both celestial (RA and Dec) and ecliptic (ecLon and ecLat) coordinates axes are shown.
    The green $+$ is the centre of the subsection.
    This sunsection was chosen because of how many asteroids there were, enough to see the shape of the box, but not so much that the lines are overwhelming to the eye.\\
    (b)
    The difference in the interpolated position of an asteroid from bulk queries to JPL Horizons, with interpolated coordinate being subtracted from the coordinate returned by Horizons. The colour bar is a different colour map to (a) as the range of time is shorter. The extent of the box is one TESS pixel, \qty{21}{\arcsec}. This is of asteroid ``1989 BC1'', but all others checked are similarly accurate (It is very expensive to check them all by brute force queries of Horizons).
  }
  \label{Fig:interpandErrPos}
\end{figure}

With TESS data coming in $\qty{10}{\minute}$ chunks, 72 interpolated points are needed between each API call.
A linear interpolation between the querried RA and Dec is preformed at the times of the TESS frames.
\texttt{TESSELLATE} saves out these time-frame pairs when doing its dataprocessing, allowing for the time coordinate of any position to be exactly at a frame time.

These interpolated positions can be seen in \autoref{Fig:interpPos} for a subsection of sector from \texttt{TESSELLATE}.
There are a few interesting features, such as the asteroids are moving in the same direction, indicated by the colouring, they come in from low RA and Dec and tend to increase both coordinates as the month of the sector progresses.
There is also a large size range in this slice of sky, ranging over \qty{5}{\mag} in absolute magnitude, which can be seen in the alpha (or transparency) of the tracks in \autoref{Fig:interpPos}.
Because of a small change in position between each frame, \citep[$\sim \qty{1}{\px}$ per \qty{30}{min} frame][]{Pal2018}, this interpolation should be accurate.

Checking against a higher frequency query to the JPL Horizons ephemeris  confirmed this accuracy on a few asteroids, as shown in \autoref{Fig:errPos}.
This is not a feasible way of getting all the positions, as the queries are per name and time, not per area and time like the cone search in Skybot.
They are also rate limited, so while a few asteroids can be checked, filling the gaps between the Skybot half day queries for all the asteroids would take far too long.
For the faster TESS FFIs, more interpolated points are needed, but a smaller the change in position between each point will be.
Forcing the interpolations to be at the same time as the TESS frames, which are known thanks to \texttt{TESSELLATE} saving them out, simplifies the matching to detections and forced photometry of these points.

\subsection{Matching to \texttt{TESSELLATE} Detections}\label{SubSec:Match}

Matching these interpolated positions to \texttt{TESSELLATE} detections is important to lower the unknown transient outputs of this pipeline.
Having interpolated their positions, the asteroids have a well sampled set of RA, Dec and time values of where they should be in the TESS data.
They should show up in a pixel for a small number of frames, of order $\sim1$.
This is the same as a lot of other transient events, a sharp rise in brightness and then disappearing quickly again.
The number of frames do change, type Ia supernova will brighten is a matter of a few hours and then dim for days, while stellar flares are of similar profile by a smaller max brightness and a correspondingly shorter decay time.

Asteroids are very short spikes, however these detection pipelines are robust.
These pipelines have already found the aforementioned supernova, stellar flares and variable stars, by matching to catalogues of those types of objects (Gaia, ZTF etc.). %? cite gaia
The job of this work is to catch all the asteroids in the set of all the transients that are returned from the pipeline.
To do this, catalogue matching is in order.

Using the \texttt{KDTree} algorithm \citep{Maneewongvatana1999} as implemented in \texttt{SciPy} \citep{2020SciPy-NMeth}, the RA and Dec coordinates of the interpolated points and the detections can be compared and matched together.
Filtering this \texttt{KDTree} output by restricting the time between spatially coincident matches to less than \qty{0.1}{\day} stops any accidental matches in position from non-asteroid detections.
A cut is then made by smallest distance, in coordinate space, and any distance must be smaller than \qty{0.01}{\degree} (\qty{36}{\arcsec}), which is within 2 TESS pixels.
Then, if multiple detection match to the same interpolated point, the detection with the smallest distance will be taken.

An example of the matches found using this \texttt{KDTree} is shown in \autoref{Fig:RADecMatch}.
It is clear that \texttt{TESSELLATE} is not detecting every frame that an asteroid is in.
Detections in \texttt{TESSELLATE} are done on a per pixel basis, and asteroids only cross each pixel for a short number of frames.
Not every interpolated point gets a match, due to a variety of reasons, which culminate in there being not an extreme enough of a difference between frames at that pixel location.
Instead, there are bunches of frequently matched points, and then patches where there are few matches detected at all.

\begin{figure}[]
  \centering
  \includegraphics[width =\textwidth]{../Test Code/Testing Figures/DetectMatchPos.pdf}
  \caption[Interpolated and Detected Positions]{The RA (right plot) and Dec (left plot) against the time of observation of the interpolated points (blue dots) or the matches to \texttt{TESSELLATE} detections (black squares). The gap at low times is from the TESS downtime in the sector, when it was passing close to earth.}
  \label{Fig:RADecMatch}
\end{figure}

\subsection{Constructing Asteroid Ligthcurves}\label{SubSec:Lightcurves}

@@

%TODO data cleaning explaination
A few steps need to be taken to clean these COM lightcurves.
Following \citet{McNeill2023}, any fluxes more than $3\sigma$ from the mean value are sigma-clipped from the forced photometry lightcurves.
The contamination from background stars the object passes is too great to allow in the lightcurve.
This does introduce gaps into an otherwise regularly sampled lightcurve, but the Lomb-Scargle periodogram is designed to deal with poorly sampled data.

The subsections made by \texttt{TESSELLATE} have a small overlap.
If an object moves between these, there will be a few frames where it is found in multiple of these subsections.
This makes the lightcurve have two simultaneous observations of the asteroid, which clearly is not possible.
At the very edge of the subsections, the asteroid can have a COM position that would have the aperture be summing over pixels that are outside the field, and thus the asteroid would have a lower flux value on the edge.
The overlap is great enough that the decrease in flux should only be happening on one of the edges at a time.
So, sorting by flux on a per asteroid basis, then removing the duplicate observation times, keeping only those with the highest flux, will clean the lightcurve.

\begin{figure}
  \centering
  \includegraphics[width =\textwidth]{../OzData/SingleBodyLCUlyssesCorrectedAgain.pdf}
  \caption[Light curves Example]{The light curves of the asteroid (5254) Ulysses throughout its entire journey across sector 29, with the flux of the interpolated positions in blue circles and the flux from the COM in orange diamonds, and the \texttt{TESSELLATE} flux for any matched points in the black squares.}
  \label{Fig:FullSingleLC}
\end{figure}

A comparison between the three light curves is interesting.
\autoref{Fig:FullSingleLC} shows these three light curves for a single body.
The COM flux seems much more stable than the interpolated positions flux on its own, and at a higher average flux count.
This should result in more accurate periodograms as there is less overall curve to the data.

The matched points are comparatively sparse for this object, which is not the case for all.
Many have no matches, and some have almost fully matched, this will be discussed more later. %ref
Lightcurves with an average flux $<10$ counts are lost in the random noise floor, so a minimum mean flux limit needs to be applied to ensure quality and reliability of any periods calculated.

\subsection{Calculation of Rotation Periods}\label{SubSec:Periods}

@@

There are a few different implementations of a Lomb-Scargle periodogram available in world of astronomy python packages.
I have used the Lomb-Scargle periodogram as implemented by \texttt{Astropy}\citep[\citet{Astropy2022} but see][for the implementation]{Vanderplas2012,Vanderplas2015}.

I first tried using the \texttt{Lightkurve} \citep{Lightkurve2018} package built for period analysis of TESS (and Kepler) data of variable stars.
\texttt{Lightkurve} was easy to use but did not allow for fine-tuning of the periodogram, or give any model of the curve it produced.
There were interesting similarities and differences between the packages.
\texttt{Lightkurve} is based on the \texttt{Astropy} methods, but does not give all the functionality, instead opting to simplify the process.
\texttt{Astropy} gives a statistic on how confident it is in the detected frequency is, the false alarm probability.

\texttt{Astropy} also returns a model, in the form of

\begin{equation}
  \label{Eq:LCModel}
  F(t;f,\vec{\theta}\,) = \theta_0 + \theta_1\cdot\sin{2\pi ft} +\theta_2\cdot\cos{2\pi ft}
\end{equation}

where the flux, $F$, is given in terms of time, $t$, the calculated frequency, $f$, and the model parameter $\vec{\theta} = [\theta_0, \theta_1,\theta_2]$.
These parameters are useful in reconstructing the sinusoid without rerunning the periodogram.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=\textwidth]{../OzData/PeriodogramUlyssesUpdated.pdf}
  \caption[Periodogram example]{\textit{Upper Panel}:The Periodogram of the COM flux of an asteroid, with the power of each frequency displayed.
    The maximum power of the periodogram is highlighted with the gold star.
    This maximum is the calculated rotation frequency.
    The known frequency, from the LCDB \citep{Warner2009}, is shown in the blue dotted lines, and a factor of 2 alias each way is shown in the purple dotted lines ($2*$) and red dotted lines ($0.5*$).
    \textit{Lower Panel}: The COM flux lightcurve and the returned model (dashed black curve) of the lightcurve from the periodogram analysis.
    The frequency of this model is the same as the gold stared point in the upper panel.
  }
  \label{Fig:PeriodEx}
\end{figure}

>>>>

\autoref{Fig:FreqVsDiam} has been recreated after data from the Light Curve Database (LCDB) \citet{Warner2009}, as cited in text. That data has been supplemented with my own work in the corresponding \autoref{Fig:FreqDiamUpdate}.

@@

Asteroids can have their rotation properties changed with time, collisions can do this, and so can light from the sun.
The YORP effect (named for the discoverers Yarkovsky, O'Keefe, Radzievskii, and Paddack) is the process of tourquing asteroids of an asymmetric shape by both bombardment of photons and thermal (re-)emission \citep{Rubincam2000}.

The data on solar system bodies will only increase with more large survey telescopes coming online soon, such as the Legacy Survey of Space and Time \citep[LSST][]{LSST2019}.
The number of new asteroids that are well characterised will increase by more than 2 orders of magnitude on the Sloan Digital Sky Survey \citep[SDSS][]{York2000} which had $\sim 88000$ objects \citet{Parker2008}.
LSST needs a dedicated pipeline for asteroid classification, the Solar System Notification Alert Processing System \citep[SNAPS][]{Trilling2023} has been developed using Zwicky Transient Facility \citep[ZTF, ][]{Bellm2018} data and can scale up handle volume of data coming from LSST.
Work is being done to understand the biases of how  asymmetric shape high amplitude variations can lead to selection effects of such large sample of asteroids \citet{Levine2023}.

Because of such a rich history of study, many asteroids have a known rotation period.
These are catalogued by \citet{Warner2009} in the Light Curve Database (LCDB).
In the October \nth{1}, 2023 release, the LCDB has nearly 9000 objects with periods determined to a quality code $U\geq 2-$, which is the minimum reliability recommended by \citeauthor{Warner2009} for statistical analysis.
This includes other TESS data from \citet{Pal2020} and \citet{Woods2021}.
The frequencies against diameter plot of these asteroids in \autoref{Fig:FreqVsDiam} shows the collection of all the asteroids in the LCDB with $U\geq 2-$.
The diameter presented in the LCDB here calculated from the absolute magnitude $H$ of the bodies, assuming an albedo.
\citeauthor{Warner2009} acknowledge that there are direct measurements of diameter for more asteroids in the present day, but continue to present the $H$ derived diameters for consistency.

The asteroid spin barrier \citep{Pravec2000} is seen in the red dashed lines in \autoref{Fig:FreqVsDiam}.
There are very few asteroids larger than $\qty{0.15}{\kilo\metre}$ rotating faster than $\sim \qty{10}{\per\day}$.
It is thought that the YORP effect, as well as collisions, spin up asteroids so fast that they pass above this barrier and then no longer have the internal strength necessary to hold themselves together.
This implies that all asteroids made to spin faster have torn themselves apart and that they have no tensile strength, i.e they are rubble piles instead of monoliths \citep{Harris1996}.
This idea of the cohesive strength of small bodies is still explored today \citep[e.g.][for V-type asteroids]{Oszkiewicz2020}.

Comparing the periods I calculate of asteroids that are already in the LCDB is key to making sure my methods are reliable, and that the periods for asteroids for which no data exists can be trusted.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\textwidth]{./Figures/Diam-FreqPlot.pdf}
  \caption[Asteroid frequencies vs diameter (LCDB)]{
    Rotation rate of asteroids by their size, recreated from the  October \nth{1}, 2023 LCDB after \citet{Warner2009}.
  }
  \label{Fig:FreqVsDiam}
\end{figure}

@@

There is ongoing analysis of subsets of the asteroid population in TESS.
\citet{Gowanlock2024} uses TESS photometry as well as  observations by ZTF to get a longer baseline on mutually observed objects while combining ground and space based observations.
There were not very many objects in both samples with known periods for comparison, with only 222 objects analysed, however they demonstrate that the technique is effective.
Fainter and unknown solar system objects can be found by shift and stacking \citep{Holman2019, Payne2019,Rice2020} or taking a fast X-ray transform \citep{Nguyen2024} of the FFI data.

This work aims to extend these studies to more sectors, hence characterising more asteroids.
The goal is to get a volume complete set of asteroid lightcurves and periods, by analysis of the whole sky as seen by TESS using data from more than just the first year of TESS's operation.
This has not been achieved, as the analysis of only one sector is presented here, but the methods are scalable.
Because of the survey properties, TESS provides a self-consistent way to measure the properties of asteroids over the full sky.
By using the higher cadence FFIs of later TESS sectors, the $P\leq\qty{3}{\hour}$ limit on the accurate periods \citep[as found by ][]{McNeill2023} can be lowered.
I also employ a different data reduction method, using reduced data from the \texttt{TESSreduce} package \citep{Ridden-Harper2021}.
This should increase the accuracy of the photometry due to the improved background subtraction process.

This work is part of a full sky transient survey using TESS, the TESS Extensive Lightcurve Logging and Analysis of Transient Events4 (\texttt{TESSELLATE}) pipeline \citep{TESSELLATE}.
To see what \texttt{TESSELLATE} detects and to help progress the pipeline, one can participate in the Cosmic Cataclysim Zooniverse project\footnote{\href{https://www.zooniverse.org/projects/cheerfuluser/cosmic-cataclysms}{https://www.zooniverse.org/projects/cheerfuluser/cosmic-cataclysms}}.
Asteroids can spike the brightness of a pixel for only a few frames while it is passing by.
This can confuse the pipeline looking for transients such as flare stars and supernova, as they can look very similar, having the brightness change in one pixel for a short time.
Knowing where the asteroids are will allow for the removal of these spikes from the transient pipeline, as well as to understand the asteroid population better.
By filtering the asteroid detections out, while also self consistently determining rotation properties of a large sample of the population, more science can be done with the one set of analyses.

@@

The \texttt{TESSELLATE} pipeline has been running on the OzSTAR supercomputing facilities.
I built this asteroid detection and subsequent lightcurve and periodogram analysis program by testing on one subsection of one CCD of one camera  of one TESS sector.
The code written can be accessed at my ASTR480 GitHub repo \footnote{\href{https://github.com/ble61/ASTR480}{https://github.com/ble61/ASTR480}}. 
Because of the large number of asteroids there are too many lightcurves to look at, so the code has to run with no human decisions. 
Once I was confident this worked as required, the same code was refactored slightly to work on OzSTAR and a large-scale analysis of a processed TESS sector was run.
% I will be looking for completeness of detections of asteroids, as well as accuracy of periods and amplitude variation.

For this work, I present analysis of TESS sector 29\footnote{\url{https://tess.mit.edu/observations/sector-29/}}, which was observed between \nth{26} August and \nth{22} September 2020.
This was the third TESS sector to be using \qty{10}{\min} cadence FFIs, which has had less analysis of the asteroids preformed.
It was chosen mainly because of this higher cadence, it was the only sector of this cadence that had been processed by \texttt{TESSELLATE} with enough time for my analysis to be preformed.
What is presented for this section is scalable, on a per-sector basis, to any sector that has been reduced.

@@

To check for asteroids in the TESS data, the positions of the asteroids with time are required.
For asteroids bright enough to be in TESS, their ephemeris is well known, so I needed to access their positions and then cross match them with transients in the TESS data.

\texttt{SkyBoT} \citep{Berthier2006} was queried to get positions of asteroids in a cone search.
A box in right ascension (RA) and  declination (Dec) space.
The search was limited to only asteroids with an average visual magnitude $V\geq \qty{20}{\mag}$.
As TESS sectors are \qty{27}{\day} long, so querying every \qty{12}{\hour} is both computationally feasible and allows for a good understanding of a bodies motion.

These positions are still very sparsely spaced in time compared to the TESS data. 
An interpolation is needed to bridge the gap, this is preformed by \texttt{numpy} \citep{Harris2020}.

Knowing where the asteroids are is helpful for finding them in the archival TESS data, but knowing more about the individual asteroids is also useful for population statistics.
The \texttt{astroquery} \citep{Ginsburg2019} package was used to query \texttt{JPL Horizons} was used to obtain the orbital elements of each asteroid, $a$, $e$, and $i$, as well as an absolute magnitude $H$, which is a good proxy for size (as discussed above). %TODO ref

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width =0.8 \textwidth]{./Figures/interpPos\_29\_1\_2\_2.pdf}
  \caption[Interpolated Positions of Asteroids]{
    The interpolated positions of asteroids in one subsection of a TESS sector.
    The colours of the lines are time sequenced, as shown in the colour bar.
    The alpha of the colours are scaled by the absolute magnitude $H$ of the asteroid.
    Both celestial coordinates (RA and Dec) and ecliptic longitude and latitude  axes are shown.
    The green $+$ is the centre of the subsection.}
    \label{Fig:interpPos}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width =0.8 \textwidth]{./Figures/1990 KC1PosCheck.pdf} %TODO one with more points???
  \caption[Interpolated Position Verification]{
  The difference in the interpolated position of an asteroid from queries to \texttt{JPL Horizons}. 
  The change in position in both arcseconds (\unit{\arcsec}) and TESS pixels are shown.  
  }\label{Fig:errPos}
\end{figure}

With TESS data coming in $\qty{10}{\minute}$ chunks, 72 interpolated points are needed between each API call.
A linear interpolation between the queried RA and Dec is preformed at the times of the TESS frames.
\texttt{TESSELLATE} saves out these time-frame pairs as part of its analysis, allowing for the time coordinate of any position to be exactly at a frame time.
This will also simplify the detection matching, as the any difference in the time will mean the detection is in a different frame.

These interpolated positions can be seen in \autoref{Fig:interpPos} for a subsection of sector from \texttt{TESSELLATE}.
There are a few interesting features, such as the asteroids are moving in the same direction, indicated by the colouring, they come in from low RA and Dec and tend to increase both coordinates as the month of the sector progresses.
There is also a large size range in this slice of sky, ranging over \qty{5}{\mag} in absolute magnitude, which can be seen in the alpha (or transparency) of the tracks in \autoref{Fig:interpPos}.

Because of a small change in position between each frame, \citep[$\sim \qty{1}{\px}$ per \qty{30}{min} frame][]{Pal2018}, this interpolation should be accurate.
Checking against a higher frequency query to the \texttt{JPL Horizons} ephemeris  confirmed this accuracy on a few asteroids, as shown in \autoref{Fig:errPos}.
This is of asteroid ``(8654) 1990 KC1'', the largest difference is only a third of a TESS pixel in RA, so the interpolations are accurate enough for these measurements. 
All others checked are similarly accurate.
This is not a feasible way of getting all the positions, as the queries are per name and time, not per area and time like the cone search in \texttt{SkyBoT}.
They are also rate limited, so while a few asteroids can be checked, filling the gaps between the \texttt{SkyBoT} half day queries for all the asteroids would be too expensive.

\subsection{Matching Asteroids to \texttt{TESSELLATE} Detections}\label{SubSec:Match}

Matching these interpolated positions to \texttt{TESSELLATE} detections is important to lower the unknown transient outputs of this pipeline.
Having interpolated their positions, the asteroids have a well sampled set of RA, Dec and time values of where they should be in the TESS data.
They should show up in a single pixel in $xy$ space for a small number of frames, of order $\sim1$.
This is the same as a lot of other transient events, a sharp rise in brightness and then disappearing quickly again.
The number of frames for events is variable, type Ia supernova will brighten is a matter of a few hours and then dim for days, while stellar flares are of similar profile by a smaller max brightness and a correspondingly shorter decay time.

Asteroids are very short events, however these detection pipelines are robust and do pick them up.
These pipelines have already found the aforementioned supernova, stellar flares and variable stars, by matching to catalogues of those types of objects (Gaia, ZTF etc.). %? cite gaia
A Goal of this work is to catch all the asteroids in the set of all the transients that are returned from the pipeline.
To do this, catalogue matching is in order.

Using the \texttt{KDTree} algorithm \citep{Maneewongvatana1999} as implemented in \texttt{SciPy} \citep{2020SciPy-NMeth}, the RA and Dec coordinates of the interpolated points and the detections can be compared and matched together.
Filtering this \texttt{KDTree} output by restricting the time between spatially coincident matches to less than \qty{0.1}{\day} stops any accidental matches in position from non-asteroid detections.
A cut is then made by distance, which must be smaller than \qty{0.01}{\degree} (\qty{36}{\arcsec}), which is within 1.7 TESS pixels.
Then, if multiple detection match to the same interpolated point, the detection with the smallest distance to the match is taken.

An example of the matches found using this \texttt{KDTree} is shown in \autoref{Fig:RADecMatch}.
It is clear that \texttt{TESSELLATE} is not detecting every frame that an asteroid is in.
Detections in \texttt{TESSELLATE} are done on a per pixel basis, and asteroids only cross each pixel for a short number of frames.
The residuals of \autoref{Fig:RADecMatch} are all within a TESS pixel, shown on the right hand axis of each of the lower pannels.

There are strings of frequently matched points, and then patches where there are few matches detected at all.
Not every interpolated point gets a match, due to a variety of reasons.
These all culminate in too small of a change in flux of the pixel from normal brightness and the frames where the asteroid is there.
The change has to be quite extreme for \texttt{TESSELLATE} to detect it.

\begin{figure}[]
  \centering
  \includegraphics[width =\textwidth]{./Figures/DetectMatchPosUlysses.pdf}
  \caption[Interpolated and Detected Positions]{The RA (right) and Dec (left) of interpolated position (blue dots) or the matches to \texttt{TESSELLATE} detections (pink hexagons) against the time of observation. The lower panels show the residuals of the difference between the positions for points that occur at the same time.}
  \label{Fig:RADecMatch}
\end{figure}

\subsection{Photometry and the Construction of Asteroid Lightcurves}\label{SubSec:Lightcurves}

@@

%TODO data cleaning explaination
A few steps need to be taken to clean these COM lightcurves.
Following \citet{McNeill2023}, any fluxes more than $3\sigma$ from the mean value are sigma-clipped recursivly from the forced photometry lightcurves, until convergence is achieved.
The contamination from background stars the object passes is too great to allow in the lightcurve.
This does introduce gaps into an otherwise regularly sampled lightcurve, but the Lomb-Scargle periodogram is designed to deal with poorly sampled data.
Lightcurves with an average flux $<10$ counts are lost in the random noise floor, so a minimum mean flux limit needs to be applied to ensure quality and reliability of any periods calculated.

The subsections of the FFIs made by \texttt{TESSELLATE} have a small overlap.
If an object moves between these, there will be a few frames where it is found in multiple of these subsections.
This makes the lightcurve have two simultaneous observations of the asteroid, which clearly is not possible.
At the very edge of the subsections, the asteroid can have a COM position that would have the aperture be summing over pixels that are outside the field, and thus the asteroid would have a lower flux value on the edge.
The overlap is great enough that the decrease in flux should only be happening on one of the edges at a time.
So, sorting by flux on a per asteroid basis, then removing the duplicate observation times, keeping only those with the highest flux, will clean the lightcurve.

\begin{figure}
  \centering
  \includegraphics[width =\textwidth]{./Figures/SingleBodyLCUlyssesgapSpan.pdf}
  \caption[Light curves Example]{The light curves of the asteroid (5254) Ulysses throughout its entire journey across sector 29. 
  The flux of the interpolated positions in blue circles,the \texttt{TESSELLATE} flux for any matched points in the pink hexagons, and the COM flux in orange diamonds.
  The mid-sector gap is highlighted by the grey box.}
  \label{Fig:FullSingleLC}
\end{figure}

A comparison between the three light curves is interesting.
\autoref{Fig:FullSingleLC} shows these three light curves for a single body.
The COM flux seems much more stable than the interpolated positions flux, and has a higher average flux count.
This should result in more accurate periodograms as there is less overall variation in the data caused by the misplaced aperture.

The matched points are comparatively sparse for this object, which is not the case for all.
Many have no matches, and some have almost fully matched, as discussed above (\autoref{SubSec:Match}) this will be discussed more later in \autoref{SubSec:Detect}.

\subsection{Calculation of Asteroid Rotation Periods}\label{SubSec:Periods}

@@

There are a few different implementations of a Lomb-Scargle periodogram available in world of astronomy python packages.
I have used the Lomb-Scargle periodogram as implemented by \texttt{Astropy}\citep[\citet{Astropy2022} but see][for the implementation]{Vanderplas2012,Vanderplas2015}.
The times are converted from MJD into seconds before the periodogram is calculated, so the frequency is calculated in \unit{\hertz}. 
The flux values have their mean subtracted so the change in flux being analysed is about 0 counts.
\texttt{Astropy} has ways of doing this internally, but I opted to do it explicitly.

\texttt{Astropy} also returns a model, in the form of
\begin{equation}
  \label{Eq:LCModel}
  F(t;f,\vec{\theta}\,) = \theta_0 + \theta_1\cdot\sin{2\pi ft} +\theta_2\cdot\cos{2\pi ft}
\end{equation}
where the flux, $F$, is given in terms of time, $t$, the calculated frequency, $f$, and the model parameter $\vec{\theta} = [\theta_0, \theta_1,\theta_2]$.
These parameters are useful in reconstructing the sinusoid without rerunning the periodogram.
\texttt{Astropy} gives a statistic on how confident it is in the detected frequency is, the false alarm probability.

@@

I originally tried using the \texttt{Lightkurve} \citep{Lightkurve2018} package built for period analysis of TESS (and Kepler) data of variable stars.
There were interesting similarities and differences between the packages.
\texttt{Lightkurve} was easy to use but did not allow for fine-tuning of the periodogram, or give any model of the best-fitting curve it produced.
\texttt{Lightkurve} is based on the \texttt{Astropy} methods, but does not give all the functionality, instead opting to simplify the process.
